First Order Alternation

### 2019-02-04 quantifiers/abstract.tex
ok

\begin{abstract}
We introduce first order alternating automata, a generalization of
boolean alternating automata, in which transition rules are described
by multisorted first order formulae, with states and internal
variables given by uninterpreted predicate terms. The model is closed
under union, intersection and complement, and its emptiness problem is
undecidable, even for the simplest data theory of equality. To cope
with this limitation, we develop an abstraction refinement
semi-algorithm based on lazy annotation of the symbolic execution
paths with interpolants, obtained by applying (i) quantifier
elimination with witness term generation and (ii) Lyndon interpolation
in the quantifier-free data theory with uninterpreted predicate
symbols. This provides a method for checking inclusion of timed and
finite-memory register automata, and emptiness of quantified predicate
automata, previously used in the verification of parameterized
concurrent programs, composed of replicated threads, with a
shared-memory communication model.
\end{abstract}

>>>>

\begin{abstract}
We introduce first order alternating automata, a generalization of
boolean alternating automata, in which transition rules are described
by multisorted first order formulae, with states and internal
variables given by uninterpreted predicate terms. The model is closed
under union, intersection and complement, and its emptiness problem is
undecidable, even for the simplest data theory of equality. To cope
with the undecidability problem, we develop an abstraction refinement
semi-algorithm based on lazy annotation of the symbolic execution
paths with interpolants, obtained by applying (i) quantifier
elimination with witness term generation and (ii) Lyndon interpolation
in the quantifier-free data theory with uninterpreted predicate
symbols. This provides a method for checking inclusion of timed and
finite-memory register automata, and emptiness of quantified predicate
automata, previously used in the verification of parameterized
concurrent programs, composed of replicated threads, with shared
memory.
\end{abstract}

### 2019-02-04 quantifiers/body.tex
ok

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{center}
\begin{minipage}{5cm}
\input ta.pdf_t
\end{minipage} 
\begin{minipage}{5cm}
\footnotesize{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]
\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata Checking Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

@@

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
cube $c$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
  is a transition rule of $\A$.
\end{compactenum}
\end{definition}
The execution semantics of automata is given in close analogy with the
classical definition of propositional alternating automata, with
transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
propositional variable and $\phi$ a positive boolean combination of
propositional variables. For instance, $q_0 \arrow{a}{} q_1 \wedge q_2
\vee q_3$ means that the automaton can choose to transition in either
both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to saying
that a transition is a \emph{minimal model} of a right hand side
formula of a rule. In our example, both $\set{q_1 \leftarrow \top, q_2
  \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow
  \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models,
however $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
  \top}$, is a model but is not minimal. The original definition of
alternating finite-state automata \cite{ChandraKozenStockmeyer81}
works around this problem by considering boolean valuations (models)
instead of formulae. In contrast, a finite description of a first
order alternating automaton must be symbolic (in terms of formulae)
because here, a given configuration can have infinitely many
successors.

An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$. We consider the
following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first order alternating automata are closed under all boolean
operations and, further, set up the ground for a semi-algorithm for
the emptiness problem.

@@

\paragraph{\em Remark}
When considering alternating executions, control-local interpolants
are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
input word. Consider, for instance, an automaton with the following
transition rules and final state $q_f$:
\[\begin{array}{rclcrclcrcl}
q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
q_1(y) & \arrow{a(x)}{} & y+x > 0 \wedge q_f & \hspace*{2mm} & 
q_1(y) & \arrow{a(x)}{} & q_1(y+x) \\ 
&&&& q_2(y) & \arrow{a(x)}{} & y-x > 0 \wedge q_f
&& q_2(y) & \arrow{a(x)}{} & q_2(y-x)
\end{array}\]
Started in an initial configuration $q_0(0)$ with an input word
$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
the automaton executes as follows:
\[\begin{array}{c}
q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots
\arrow{(a,\nu_{n-1})}{} 
\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
\arrow{(a,\nu_n)}{} \emptyset
\end{array}\]
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 = 0$. Note that a control-local formula
using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. \hfill$\blacksquare$

@@

\begin{proposition}\label{prop:local-gli}
  If there exists a Lyndon interpolant for any two formulae $\phi$ and
  $\psi$, in the first order theory of data with uninterpreted
  predicate symbols, such that $\phi \wedge \psi$ is unsatisfiable,
  then any sequence of input events $\alpha = a_1 \ldots a_n \in
  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, has a
  local GLI $(I_0,\ldots,I_n)$.
\end{proposition}
\proof{ By definition, $\accform{\alpha}$ is the formula:
  \[\stamp{\iota}{0} \wedge 
 \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
 \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
 We define the formulae: 
 \[\begin{array}{rcl}
 \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
 \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 \end{array}\]
 Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
 $\voc{\varphi_i} \subseteq
 \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
       Lyndon's Interpolation Theorem for the formulae
       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed}

@@

\begin{example}[Contd. from Examples \ref{ex:quant-elim} and
    \ref{ex:pred-elim}]\label{ex:gli} The formula $\substform{\alpha}$
  (Example \ref{ex:pred-elim}) is unsatisfiable and let $\tau_2 \isdef
  z_1$ be the witness term for the universally quantified variable
  $z_2$. Replacing $z_2$ with $\tau_2$ in the matrix of
  $\quantform{\alpha}$ (Example \ref{ex:quant-elim}) yields the
  unsatisfiable conjunction below, obtained after trivial
  simplifications:
\[\begin{array}{lcll}
           [z_1 \geq 0 \wedge \stamp{q}{0}(z_1)] & \wedge &
           [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)] ~\wedge \\
           && [\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)]
\end{array}\]
A non-local GLI for the above conjunction is the sequence of formulae: 
\[(\stamp{q}{0}(z_1) \wedge z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
\wedge z_1\geq0,~\bot) \hspace*{4cm}\blacksquare\]
\end{example}

@@

\begin{table}[htb]
\vspace*{-\baselineskip}
\begin{center}
{\fontsize{7}{8}\selectfont
\begin{tabular}{||l|c|c|c|c|c||}
\hline
Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
\hline
incdec.pa & 499 & no & 21 & 17 & 779 \\
\hline
localdec.pa & 678 & no & 49 & 35 & 1814 \\
\hline
ticket.pa & 4250 & no & 229 & 91 & 9543 \\
\hline
count\_thread0.pa & 9767 & no & 154 & 128 & 8553 \\
\hline
count\_thread1.pa & 10925 & no & 766 & 692 & 76771 \\
\hline
local0.pa & 10595 & no & 73 & 27 & 1431 \\
\hline
local1.pa & 11385 & no & 1135 & 858 & 101042 \\
\hline
array\_rotation.ada & 1834 & yes & 9 & 8 & 1543 \\
\hline
array\_simple.ada & 3440 & yes & 11 & 10 & 6787 \\
\hline
array\_shift.ada & 874 & yes & 6 & 5 & 413 \\
\hline
abp.ada & 6909 & no & 52 & 47 & 4788 \\
\hline
train.ada & 1823 & yes & 68 & 67 & 7319 \\
\hline
hw1.ada & 322 & Solver Error & / & / & / \\
\hline
hw2.ada & 674 & yes & 20 & 22 & 4974 \\
\hline
rr-crossing.foada & 1780 & yes & 67 & 67 & 7574 \\
% \hline
% running2.foada & 1217 & yes & 13 & 13 & 788 \\
\hline
train-simple1.foada & 5421 & yes & 43 & 44 & 2893 \\
\hline
train-simple2.foada & 10177 & yes & 111 & 113 & 8386 \\
\hline
train-simple3.foada & 15961 & yes & 196 & 200 & 15041 \\
% \hline
% fischer-mutex1.foada & 1566 & no & 31 & 31 & 1552 \\
\hline
fischer-mutex2.foada & 3000 & yes & 23 & 23 & 808 \\
\hline
fischer-mutex3.foada & 4452 & yes & 33 & 33 & 1154 \\
\hline
\end{tabular}
}
\caption{Experiments with First Order Alternating Automata}
\label{tab:experiments}
\end{center}
\vspace*{-2\baselineskip}
\end{table}

>>>>

\begin{figure}[htb]
\vspace*{-2\baselineskip}
\begin{center}
\begin{minipage}{5cm}
\input ta.pdf_t
\end{minipage} 
\begin{minipage}{5cm}
\footnotesize{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]
\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata Checking Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

@@

\begin{definition}\label{def:execution}
Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
cube $c$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
labeled with configurations, such that:
\begin{compactenum}
\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
  configurations labeling the roots of $T_1,T_2,\ldots$ and
%
\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
  is a transition rule of $\A$.
\end{compactenum}
\end{definition}

The execution semantics of automata is given in close analogy with the
classical definition of propositional alternating automata, with
transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
propositional variable and $\phi$ a positive boolean combination of
propositional variables. For instance, $q_0 \arrow{a}{} (q_1 \wedge
q_2) \vee q_3$ means that the automaton can choose to transition in
either both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to
saying that a transition is a \emph{minimal model} of a right hand
side formula of a rule. In our example, both $\set{q_1 \leftarrow
  \top, q_2 \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1
  \leftarrow \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are
minimal models, however $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \top}$, is a model but is not minimal. The
original definition of alternating finite-state automata
\cite{ChandraKozenStockmeyer81} works around this problem by
considering boolean valuations (models) instead of formulae. In
contrast, a finite description of a first order alternating automaton
must be symbolic (in terms of formulae) because here, a given
configuration can have infinitely many successors.

An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$.

\begin{example}\label{ex:foada}
  Consider the automaton $\A = \tuple{\set{a}, \set{x},
    \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
  consists of the rules below:
  \[\begin{array}{rclcrclcrcl}
  q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} &
  q_1(y) & \arrow{a(x)}{} & q_1(y+x) \hspace*{2mm} &
  q_1(y) & \arrow{a(x)}{} & y > 0 \wedge q_f \\ 
  &&&& q_2(y) & \arrow{a(x)}{} & q_2(y-x) &
  q_2(y) & \arrow{a(x)}{} & y > 0 \wedge q_f
  \end{array}\]
  A possible execution tree of this automaton is the following:
  \begin{center}
    \input exec.pdf_t
  \end{center}
  The execution is not accepting, since its frontier is not labeled
  with final configurations everywhere. It is not hard to see that
  this automaton has no accepting execution. \hfill$\blacksquare$
\end{example}

In the rest of this paper, we are concerned with the following
problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first order alternating automata are closed under all boolean
operations and, further, set up the ground for a semi-algorithm for
the emptiness problem.

@@

\paragraph{\em Remark}
When considering alternating executions, control-local interpolants
are not always enough to prove emptiness, because of the
synchronization of several branches of the computation on the same
input word. Consider, for instance, the automaton from Example
\ref{ex:foada}, started in an initial configuration $q_0(0)$, on an
input word $(a,\nu_1) \ldots (a,\nu_n)$, such that $\nu_i(x)=k_i$, for
all $i \in [1,n]$, the automaton executes as follows:
\[\begin{array}{c}
\set{(q_0,0)} \arrow{(a,\nu_1)}{} \set{(q_1,k_1),(q_2,-k_1)} \ldots
\arrow{(a,\nu_{n})}{} 
\set{(q_1,\sum_{i=1}^{n} k_i),(q_2,-\sum_{i=1}^{n} k_i)}
\end{array}\]
An overapproximation of the set of cubes generated after one or more
steps is defined by the formula $\exists x_1 \exists x_2 ~.~ q_1(x_1)
\wedge q_2(x_2) \wedge x_1+x_2 \geq 0$. Note that a control-local formula
using one occurrence of a predicate would give a too rough
overapproximation of this set, unable to prove the emptiness of the
automaton. This is because any formula with a single occurrence of an
atom of the form $q(x)$ would lose the constraint $x_1+x_2 \geq 0$
between the values occurring at the same level, on both branches of
the execution. \hfill$\blacksquare$

@@

\begin{proposition}\label{prop:local-gli}
  If there exists a Lyndon interpolant for any two formulae $\phi$ and
  $\psi$, in the first order theory of data with uninterpreted
  predicate symbols, such that $\phi \wedge \psi$ is unsatisfiable,
  then any sequence of input events $\alpha = a_1 \ldots a_n \in
  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, has a
  local GLI $(I_0,\ldots,I_n)$.
\end{proposition}
\paragraph{\em Proof.} By definition, $\accform{\alpha}$ is the formula:
\[\begin{array}{c}
\stamp{\iota}{0} \wedge 
 \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
 \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 \end{array}\]
 We define the formulae: 
 \[\begin{array}{rcl}
 \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
 \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
 ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
 \end{array}\]
 Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
 $\voc{\varphi_i} \subseteq
 \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
       Lyndon's Interpolation Theorem for the formulae
       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed

@@

\begin{example}[Contd. from Examples \ref{ex:quant-elim} and
    \ref{ex:pred-elim}]\label{ex:gli} The formula $\substform{\alpha}$
  (Example \ref{ex:pred-elim}) is unsatisfiable and let $\tau_2 \isdef
  z_1$ be the witness term for the universally quantified variable
  $z_2$. Replacing $z_2$ with $\tau_2$ ($z_1$) in the matrix of
  $\quantform{\alpha}$ (Example \ref{ex:quant-elim}) yields the
  unsatisfiable conjunction below, obtained after trivial
  simplifications:
\[\begin{array}{lcll}
           [z_1 \geq 0 \wedge \stamp{q}{0}(z_1)] & \wedge &
           [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)] ~\wedge \\
           && [\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)]
\end{array}\]
A non-local GLI for the above conjunction is the sequence of formulae: 
\[(\stamp{q}{0}(z_1) \wedge z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
\wedge z_1\geq0,~\bot) \hspace*{4cm}\blacksquare\]
\end{example}

@@

\begin{table}[htb]
%\vspace*{-\baselineskip}
\begin{center}
{\fontsize{7}{8}\selectfont
\begin{tabular}{||l|c|c|c|c|c||}
\hline
Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
\hline
incdec.pa & 499 & no & 21 & 17 & 779 \\
\hline
localdec.pa & 678 & no & 49 & 35 & 1814 \\
\hline
ticket.pa & 4250 & no & 229 & 91 & 9543 \\
\hline
count\_thread0.pa & 9767 & no & 154 & 128 & 8553 \\
\hline
count\_thread1.pa & 10925 & no & 766 & 692 & 76771 \\
\hline
local0.pa & 10595 & no & 73 & 27 & 1431 \\
\hline
local1.pa & 11385 & no & 1135 & 858 & 101042 \\
\hline
array\_rotation.ada & 1834 & yes & 9 & 8 & 1543 \\
\hline
array\_simple.ada & 3440 & yes & 11 & 10 & 6787 \\
\hline
array\_shift.ada & 874 & yes & 6 & 5 & 413 \\
\hline
abp.ada & 6909 & no & 52 & 47 & 4788 \\
\hline
train.ada & 1823 & yes & 68 & 67 & 7319 \\
\hline
hw1.ada & 322 & Solver Error & / & / & / \\
\hline
hw2.ada & 674 & yes & 20 & 22 & 4974 \\
\hline
rr-crossing.foada & 1780 & yes & 67 & 67 & 7574 \\
% \hline
% running2.foada & 1217 & yes & 13 & 13 & 788 \\
\hline
train-simple1.foada & 5421 & yes & 43 & 44 & 2893 \\
\hline
train-simple2.foada & 10177 & yes & 111 & 113 & 8386 \\
\hline
train-simple3.foada & 15961 & yes & 196 & 200 & 15041 \\
% \hline
% fischer-mutex1.foada & 1566 & no & 31 & 31 & 1552 \\
\hline
fischer-mutex2.foada & 3000 & yes & 23 & 23 & 808 \\
\hline
fischer-mutex3.foada & 4452 & yes & 33 & 33 & 1154 \\
\hline
\end{tabular}
}
\caption{Experiments with First Order Alternating Automata}
\label{tab:experiments}
\end{center}
\vspace*{-2\baselineskip}
\end{table}
